---
title: "Getting Started with orthomr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with orthomr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(orthomr)
library(polynom)
library(dplyr)
```

## Introduction

The `orthomr` package provides tools for non-linear Mendelian randomization using orthogonal polynomials in a multivariable MR context. This vignette demonstrates a complete workflow from simulation to analysis.

## Simulate Data

First, we simulate genetic data with a non-linear relationship between exposure and outcome:

```{r simulate}
simulation_result <- sim_exp_out_non_linear(
  include_pleiotropy = TRUE,
  sample_size = 10000,
  n_instruments = 30,
  exposure_to_outcome_func = as.function(polynomial(coef = c(1, 10, 3, 0))),
  confounder_to_exposure_func = as.function(polynomial(coef = c(3, 9, -6, 1))),
  confounder_to_outcome_func = as.function(polynomial(coef = c(3, -2, 4)))
)
```

Extract the simulated data:

```{r extract-data}
outcome_observed <- simulation_result[["simulated_data"]]$outcome_observed
exposure_observed <- simulation_result[["simulated_data"]]$exposure_observed
outcome_true <- simulation_result[["simulated_data"]]$outcome_true
exposure_true <- simulation_result[["simulated_data"]]$exposure_true
genotypes <- simulation_result[["genotypes"]]
allele_frequencies <- simulation_result[["allele_frequencies"]]
```

## Create Orthogonal Polynomials

We create orthogonal polynomials up to degree 5:

```{r orthogonal-polynomials}
n_degrees <- 5
poly_result <- orthopol(degree = n_degrees, exposure_values = exposure_observed)

# View structure
str(poly_result, max.level = 1)
```

## Perform GWAS

### GWAS for Exposures (Polynomial Terms)

```{r gwas-exposures}
gwas_exposure_list <- list()
for (j in 1:n_degrees) {
  gwas_temp <- c()
  for (i in 1:ncol(genotypes)) {
    model <- lm(poly_result[["values"]][, j] ~ genotypes[, i])
    gwas_temp <- data.frame(
      ins = paste0("X", i),
      beta = coefficients(model)[2],
      se = coef(summary(model))[, "Std. Error"][2],
      pval = coef(summary(model))[, "Pr(>|t|)"][2],
      af = allele_frequencies[i]
    ) %>%
      bind_rows(gwas_temp, .)
  }
  gwas_exposure_list[[j]] <- gwas_temp
}
```

### GWAS for Outcome

```{r gwas-outcome}
gwas_outcome <- c()
for (i in 1:ncol(genotypes)) {
  model <- lm(outcome_observed ~ genotypes[, i])
  gwas_outcome <- data.frame(
    ins = paste0("X", i),
    beta = coefficients(model)[2],
    se = coef(summary(model))[, "Std. Error"][2],
    pval = coef(summary(model))[, "Pr(>|t|)"][2],
    af = allele_frequencies[i]
  ) %>%
    bind_rows(gwas_outcome, .)
}
```

## Format Data for MVMR

Merge instruments and format for TwoSampleMR:

```{r format-mvmr}
mvmr_exposure_data <- merge_exp_inst(gwas_exposure_list)

mvmr_outcome_data <- gwas_outcome %>%
  filter(ins %in% mvmr_exposure_data$SNP) %>%
  rename(SNP = ins) %>%
  mutate(outcome = "outcome") %>%
  mutate(id.outcome = "outcome") %>%
  mutate(effect_allele.outcome = "A") %>%
  mutate(other_allele.outcome = "C") %>%
  rename(beta.outcome = beta) %>%
  rename(se.outcome = se) %>%
  rename(pval.outcome = pval) %>%
  rename(eaf.outcome = af)
```

## Run Complete MR Analysis

The `mr_sim_res()` function performs the complete workflow:

```{r run-analysis}
analysis_results <- mr_sim_res(
  mvmr_exposure_data = mvmr_exposure_data,
  mvmr_outcome_data = mvmr_outcome_data,
  simulation_data = simulation_result,
  exposure_values = NULL,
  inner_interval = 0.90,
  include_intercept = TRUE,
  orthogonal_poly_coef = poly_result$coefficients
)
```

## View Results

### MVMR Results

```{r view-mvmr-results}
analysis_results$mvmr_results
```

### Instrument Strength

```{r view-instrument-strength}
analysis_results$instrument_strength
```

### Visualization

```{r view-plots, fig.width=10, fig.height=5}
analysis_results$plots
```

### Pairwise Comparisons

```{r view-pairwise-full, fig.width=10, fig.height=8}
analysis_results$pairwise_comparison_full
```

```{r view-pairwise-inner, fig.width=10, fig.height=8}
analysis_results$pairwise_comparison_inner
```

## Conclusion

This workflow demonstrates how to:

1. Simulate data with non-linear relationships
2. Create orthogonal polynomials
3. Perform GWAS for multiple polynomial degrees
4. Run multivariable Mendelian randomization
5. Visualize and interpret results

For more details on individual functions, see their respective help pages (e.g., `?orthopol`, `?mr_sim_res`).
